Komols Complexity, Coresets, and Sketches in Machine Learning

### 2018-12-28 komlos.tex

\section{Universal Coresets for Analytical Functions}

We provide a coreset suitable for analytical functions of the inner product $\ip{q,x}$. The idea is to find a set of signs that simultaneously balance $\ip{q,x}^k$ for all powers $k$ and unit vectors $q$. By controlling all powers of $\ip{q,x}$ we control any sum of these powers. It follows that this coreset can be used to control, for example, the logistic loss function $L(q,x) = \log(1+\exp(\ip{q,x}))$.

We start some notation and trivial properties. For a vector $q \in \R^d$ let $q^{\otimes k}$ to represent the $k$-dimensional tensor obtained from the outer product of $q$ with itself in all the dimensions. For a $k$ dimensional tensor with $d^k$ entries $X$ we consider the measure
$\|X\|_{T_k} = \max_{q \in \R^d, \|q\|=1} \left| \langle X, q^{\otimes k}\rangle \right|$.
\begin{lemma}
$\|X\|_{T_k}$ is a norm
\end{lemma}
\begin{proof}
We prove the claim directly from the definition of a norm.
Notice that for any $X \neq 0$, $\ip{X, q^{\otimes}}$ is a non-zero polynomial in $q$. It follows that there must be $q$ for which it's value is non-zero, meaning that $\|X\|_{T_k}=0$ iff $X=0$. For a scalar $a$, we clearly have by definition that
$$ \|aX\|_{T_k} = |a|\|X\|_{T_k} $$
Lastly, by the max definition we clearly have
$$ \|X+Y\|_{T_k} =  \max_{q \in \R^d, \|q\|=1} \left| \langle X+Y, q^{\otimes k}\rangle \right| \leq $$
$$\max_{q \in \R^d, \|q\|=1} \left| \langle X, q^{\otimes k}\rangle \right| + \max_{q \in \R^d, \|q\|=1} \left| \langle Y, q^{\otimes k}\rangle \right| = \|X\|_{T_k} + \|Y\|_{T_k}$$
\end{proof}

@@

\begin{lemma}\label{uc}
For any set of vectors $x_i \in \R^d$ there exist a set of signs $\sigma_i$ such that for all $k$ simultaneously $\left\| \sum_i \sigma_i x_i^{\otimes k} \right\|_{T_k} \le O(\sqrt{d k\log^{3}{k}})$ (the $3$ power of the term $\log(k)$ can be reduced to any constant power larger than $2$). 
\end{lemma}
\begin{proof}
The proof will use Banaszczyk's theorem. 
Let $\mathcal K$ be a convex body in Euclidean space with Gaussian measure at least 1/2 ($\Pr[g \in \mathcal K] \ge 1/2$ where $g$ is i.i.d.\ Gaussian).
Let $x_1,\ldots,x_n$ be vectors with $\|x_i\| =O(1)$. 
Then, there exist signs $\sigma_i$ such that $\sum \sigma_i x_i \in C \mathcal K$ for some constant $C$.

@@

We now need to show that the Gaussian measure of $\mathcal K$ is at least $1/2$. 
That is, with probability at least $1/2$ a vector of random Gaussian entrees $g$ belongs to $\mathcal K$.
Considering a random i.i.d.\ Gaussian Tensor $g_k \in \R^{d^k}$. 
A trivial modification of Theorem 1 from \cite{tomioka2014spectral} shows that $\Pr[\|g_k\|_{T_k} \ge c\sqrt{d\log(k)}] \le 1/10k^2$ for some constant $c$. The only change needed in the proof is the size of the epsilon net which changes from $(2\log(3/2)/k)^{kd}$ for \cite{tomioka2014spectral} to $(2\log(3/2)/k)^d$, where the reason we require a net over a smaller space is due to us bounding the inner product with a rank one tensor rather than rank $k$. Union bounding on all values of $k$ we get $\sum_k 1/10k^2 \le 1/2$ which shows $g = [g_1, \operatorname{flat}(g_2), \operatorname{flat}(g_3), \ldots]$ belongs to $\mathcal K$ with probability al least $1/2$. The function $\operatorname{flat}(X)$ simply strings the entry values of the tensor $X$ into a vector.

We now define a mapping $\psi(x)$ of $x\in \R^d$ to a high dimensional space. 
$$\psi(x) = [x, \frac{\operatorname{flat}(x^{\otimes 2})}{\sqrt{2\log^2(2)}}, \frac{\operatorname{flat}(x^{\otimes 3})}{\sqrt{3\log^2(3)}}, \ldots,\frac{\operatorname{flat}(x^{\otimes k})}{\sqrt{k\log^2(k)}},\ldots]$$
Finally, note that for $\|x\| \le 1$ we have $\|\psi(x)\|_2 = (\sum_k  1/k\log^2(k))^{1/2} = O(1)$.

We are now ready to apply Banaszczyk's theorem. 
There exist signs $\sigma_i$ such that $\psi  = \sum_i \sigma_i \psi(x_i) \in C \mathcal K$.
Since $\psi_k = \sum_i \sigma_i x_i^{\otimes k}/\sqrt{k \log^2{k}}$ we get that $\max_k \|\sum_i \sigma_i  x_i^{\otimes k}\|_{T_k} \le \sqrt{d k \log^{3}(k)}$

>>>>

\section{Universal Coresets for Analytic Functions of Dot Products}

%<<<<<<< HEAD
%\begin{lemma}\label{uc}
%For any set of vectors $x_i \in \R^d$ there exist a set of signs $\sigma_i$ such that for all $k$ simultaneously $\|\sum_i \sigma_i x_i^{\otimes k}\| \le O(\sqrt{d k\log^{3}{k}})$ (the $3$ power of the term $\log(k)$ can be reduced to any constant power larger than $2$). Here 
%$\|X\| = \max_{q, \|q\|=1} \langle X, q^{\otimes k}\rangle$.
%=======
We provide a coreset suitable for analytical functions of the inner product $\ip{q,x}$. The idea is to find a set of signs that simultaneously balance $\ip{q,x}^k$ for all powers $k$ and unit vectors $q$. By controlling all powers of $\ip{q,x}$ we control any sum of these powers. It follows that this coreset can be used to control, for example, the logistic loss function $L(q,x) = \log(1+\exp(\ip{q,x}))$.

We start with some notation and trivial properties. 
For a vector $q \in \R^d$ let $q^{\otimes k}$ represent the $k$-dimensional tensor obtained from the outer product of $q$ with itself $k$ times. For a $k$ dimensional tensor with $d^k$ entries $X$ we consider the measure
$\|X\|_{T_k} = \max_{q \in \R^d, \|q\|=1} \left| \langle X, q^{\otimes k}\rangle \right|$.
\begin{fact}
$\|X\|_{T_k}$ is a norm
\end{fact}
\begin{proof}
We prove the claim directly from the definition of a norm.
Notice that for any $X \neq 0$, $\ip{X, q^{\otimes}}$ is a non-zero polynomial in $q$. It follows that there must be $q$ for which its value is non-zero, meaning that $\|X\|_{T_k}=0$ iff $X=0$. For a scalar $a$, we clearly have by definition that
$\|aX\|_{T_k} = |a|\|X\|_{T_k}$.  Lastly, by the max definition we clearly have
$ \|X+Y\|_{T_k} =  \max_q \left| \langle X+Y, q^{\otimes k}\rangle \right| \leq 
\max_q \left| \langle X, q^{\otimes k}\rangle \right| + \max_q\left| \langle Y, q^{\otimes k}\rangle \right| = \|X\|_{T_k} + \|Y\|_{T_k}$
\end{proof}
%>>>>>>> 56549e52bc0636f8545357647f053d28b3c0c7df

@@

\begin{lemma}\label{uc}
For any set of vectors $x_i \in \R^d$ there exist a set of signs $\sigma_i$ such that for all $k$ simultaneously $\left\| \sum_i \sigma_i x_i^{\otimes k} \right\|_{T_k} \le O(\sqrt{d k\log^{3}{k}})$ (the $3$ power of the term $\log(k)$ can be reduced to any constant power larger than $2$). 
\end{lemma}
\begin{proof}
The proof will use Banaszczyk's theorem. 
Let $\mathcal K$ be a convex body in Euclidean space with Gaussian measure at least 1/2 ($\Pr[g \in \mathcal K] \ge 1/2$ when $g$ is i.i.d.\ Gaussian).
Let $x_1,\ldots,x_n$ be vectors with $\|x_i\| =O(1)$. 
Then, there exist signs $\sigma_i$ such that $\sum \sigma_i x_i \in C \mathcal K$ for some constant $C$.

@@

We now need to show that the Gaussian measure of $\mathcal K$ is at least $1/2$. 
That is, with probability at least $1/2$ a vector of random Gaussian entrees $g$ belongs to $\mathcal K$.
Consider a random i.i.d.\ Gaussian Tensor $g_k \in \R^{d^k}$. 
%<<<<<<< HEAD
%A trivial modification of Theorem 1 from \cite{tomioka2014spectral} shows that $\Pr[\|g_k\| \ge c\sqrt{d\log(k)}] \le 1/10k^2$ for some constant $c$. The only change needed in the proof is the size of the epsilon net which changes from $(2\log(3/2)/k)^{kd}$ for \cite{tomioka2014spectral} to $(2\log(3/2)/k)^d$. Union bounding on all values of $k$ we get $\sum_k 1/10k^2 \le 1/2$ which shows $g = [g_1, \operatorname{flat}(g_2), \operatorname{flat}(g_3), \ldots]$ belongs to $\mathcal K$ with probability al least $1/2$. The function $\operatorname{flat}(X)$ simply strings the entry values of the tensor $X$ into a vector.
%%
%We now define a mapping $\psi(x)$ of $x\in \R^d$ to a high dimensional space. 
%$$\psi(x) = [x, \frac{\operatorname{flat}(x^{\otimes 2})}{\sqrt{2\log^2(2)}}, \frac{\operatorname{flat}(x^{\otimes 3})}{\sqrt{3\log^2(3)}}, \ldots,\frac{\operatorname{flat}(x^{\otimes k})}{\sqrt{k\log^2(k)}},\ldots]$$
%Note that for $\|x\| \le 1$ we have $\|\psi(x)\| = (\sum_k  1/k\log^2(k))^{1/2} = O(1)$.
%
%=======
A trivial modification of Theorem 1 from \cite{tomioka2014spectral} shows that $\Pr[\|g_k\|_{T_k} \ge c\sqrt{d\log(k)}] \le 1/10k^2$ for some constant $c$. The only change needed in the proof is the size of the epsilon net which changes from $(2\log(3/2)/k)^{kd}$ for \cite{tomioka2014spectral} to $(2\log(3/2)/k)^d$, where the reason we require a net over a smaller space is due to us bounding the inner product with a rank one tensor rather than rank $k$. Union bounding on all values of $k$ we get $\sum_k 1/10k^2 \le 1/2$ which shows $g = [g_1, \operatorname{flat}(g_2), \operatorname{flat}(g_3), \ldots]$ belongs to $\mathcal K$ with probability at least $1/2$.

We now define a mapping $\psi(x)$ of $x\in \R^d$ to a high dimensional space. The function $\operatorname{flat}(X)$ simply strings the entry values of the tensor $X$ into a vector.
$$\psi(x) = [x, \frac{\operatorname{flat}(x^{\otimes 2})}{\sqrt{2\log^2(2)}}, \frac{\operatorname{flat}(x^{\otimes 3})}{\sqrt{3\log^2(3)}}, \ldots,\frac{\operatorname{flat}(x^{\otimes k})}{\sqrt{k\log^2(k)}},\ldots]$$
Finally, note that for $\|x\| \le 1$ we have $\|\psi(x)\|_2 = (\sum_k  1/k\log^2(k))^{1/2} = O(1)$.

%>>>>>>> 56549e52bc0636f8545357647f053d28b3c0c7df
We are now ready to apply Banaszczyk's theorem. 
There exist signs $\sigma_i$ such that $\psi  = \sum_i \sigma_i \psi(x_i) \in C \mathcal K$.
Since $\psi_k = \sum_i \sigma_i x_i^{\otimes k}/\sqrt{k \log^2{k}}$ we get that $\max_k \|\sum_i \sigma_i  x_i^{\otimes k}\|_{T_k} \le \sqrt{d k \log^{3}(k)}$
